---
title: "Cleaning and Wrangling Data"
author: "Michelle Mohr"
date: "2023-04-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load packages
```{r load packages, warning=FALSE, message=FALSE} 
library(dplyr)
library(tidyr)
library(readr)


```


# Introduction
About the data: 
This data was downloaded from [this link](https://knb.ecoinformatics.org/view/df35b.304.2), on April 4, 2023.

## Read data

```{r read data, warning=FALSE, message=FALSE}
catch_original <-read_csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1")
```

## Explore data

```{r explore, eval=FALSE, echo=FALSE}

colnames(catch_original)

head(catch_original)

summary(catch_original)

sort(unique(catch_original$Year))

unique(catch_original$Region)

```


# Clean and Reshape data

## Selecting and Removing columns with `select()`

```{r select}

catch_data <- catch_original %>% 
  select(Region, Year, Chinook, Sockeye, Coho, Pink, Chum)

head(catch_data)


catch_data <- catch_original %>% 
  select(-All, - notesRegCode)

head(catch_data)

```


## Check columns types 

```{r checking}

glimpse(catch_data) 

```
## Changing column content using `mutate()`

```{r changing data type}

catch_clean <- catch_data %>% 
  mutate(Chinook = as.numeric(Chinook))

head(catch_clean)

```

## Checking NA

```{r checking na}

i <- which(is.na(catch_clean$Chinook))

catch_data[i,]
```

## Fixing Chinook column
```{r}
catch_clean <- catch_data %>% 
  mutate(Chinook = if_else(Chinook =="I",
                           "1", 
                           Chinook),
         Chinook = as.numeric(Chinook))

## Check
catch_clean[i,]
```

## Reshape data

From wide to long use `pivot_longer()`

```{r reshape}

catch_long <- catch_clean %>% 
  pivot_longer(cols = - c(Region, Year), 
               names_to = "species", 
               values_to = "catch")
head(catch_long)

```

From long to wide use `pivot_wider()`

```{r reshape wide}

catch_wide <- catch_long %>% 
  pivot_wider(names_from = species, 
              values_from = catch)

head(catch_wide)
```

## Renaming column 

```{r}
catch_long <- catch_long %>% 
  rename(catch_thousand = catch)

colnames(catch_long)
```
## Add a column using `mutate()`

```{r adding column}

catch_long <- catch_long %>% 
  mutate(catch = catch_thousand*1000) %>% 
  select(-catch_thousand)

head(catch_long)
```

# Split-Apply-Combine workflow
We are going to use the `group_by()` and summarize()`

```{r}

mean_region <- catch_long %>% 
  group_by(Region) %>% 
  summarize(catch_mean = mean(catch))

head(mean_region)

mean_reg_yr <- catch_long %>% 
  group_by(Region) %>% 
  summarize(catch_mean = mean(catch))

head(mean_reg_yr)

n_region <- catch_long %>% 
  group_by(Region) %>% 
  summarize(obs_n = n())

head(n_region)


```


## Filtering by rows with `filter()`

```{r filtering}

sse_catch <- catch_long %>% 
  filter(Region == "SSE")

head(sse_catch)
```


