---
title: 'R Practice: Tidy Data and Joins'
author: "Michelle Mohr"
date: "2023-04-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Load Libraries
```{r libraries}
library(readr)
library(dplyr)
library(lubridate)
```

# Download Data
```{r data}
read_csv("https://portal.edirepository.org/nis/dataviewer?packageid=knb-lter-cap.256.10&entityid=53edaa7a0e083013d9bf20322db1780e")
read_csv("https://portal.edirepository.org/nis/dataviewer?packageid=knb-lter-cap.256.10&entityid=b2466fa5cb5ed7ee1ea91398fc291c59")
read_csv("https://portal.edirepository.org/nis/dataviewer?packageid=knb-lter-cap.256.10&entityid=81bf72420e69077097fb0790dcdc63a6")
read_csv("https://portal.edirepository.org/nis/dataviewer?packageid=knb-lter-cap.256.10&entityid=58f863b7e3066e68536a9cacdc7bd58e")
```
# Read in data
```{r read in data}
bird_observations <- read_csv("https://portal.edirepository.org/nis/dataviewer?packageid=knb-lter-cap.256.10&entityid=53edaa7a0e083013d9bf20322db1780e")
surveys <- read_csv("https://portal.edirepository.org/nis/dataviewer?packageid=knb-lter-cap.256.10&entityid=b2466fa5cb5ed7ee1ea91398fc291c59")
sites <- read_csv("https://portal.edirepository.org/nis/dataviewer?packageid=knb-lter-cap.256.10&entityid=81bf72420e69077097fb0790dcdc63a6")
taxalist <- read_csv("https://portal.edirepository.org/nis/dataviewer?packageid=knb-lter-cap.256.10&entityid=58f863b7e3066e68536a9cacdc7bd58e")
```
# View data
```{r view data}
# view data in a spreadsheet-style viewer
View(bird_observations)

# returns dimensions of the dataframe by number of rows and number of cols
dim(bird_observations)

# returns the top six rows of the dataframe
head(bird_observations)

# returns all the columns and some info about the cols
glimpse(bird_observations)

# similar to glimpse but returns some summary statistics about the cols
summary(bird_observations)

# returns column names 
names(bird_observations)

# returns unique values in a column. In this case we can see all the different bird species IDs
unique(bird_observations$species_id)
```
## Create a subset of bird observations 
```{r subset bird observations}
birds_subset <- bird_observations %>% 
  filter(species_id %in% c("BHCO", "RWBL")) %>% 
  filter(site_id %in% c("LI-W", "NU-C"))
```
## Joining Data 
### Left join to merge birds subset with sites and taxalist tables 
```{r left join}
# syntax using pipe
birds_left <- birds_subset %>% 
  left_join(y = sites, by = "site_id")
# don't see x = birds_subset here because piping in birds_subset means it automatically assumes birds_subset as x.

# syntax without pipe
birds_left <- left_join(x = birds_subset, y = sites, by = "site_id")
```

### Full join to merge birds subset and sites tables 
```{r full join}
birds_full <- birds_subset %>% 
  full_join(y = sites, by = "site_id")
```

### Inner join to merge birds subset and taxalist data
```{r inner join}
birds_inner <- birds_subset %>% 
  inner_join(y = taxalist, by = "species_id")
```

## Wrangle and Joing Data 
### Wrangle bird observations data and merge the data with all the other data tables 
```{r wrangle and merge}
bird_observations <- bird_observations %>%  
  rename(bird_obs_notes = notes)
```

### Create a subset that contains all observations in the bird observations data frames 
```{r subset of all observations in bird observations}
birds_obs_subset <- bird_observations %>% 
full_join(y = taxalist, by = "species_id") %>% 
full_join(y = sites, by = "site_id") %>% 
full_join(y = surveys, by = c("site_id", "survey_id")) %>% 
  select(survey_date, common_name, park_name, bird_count, observer)

```

