---
title: "RMarkdown Analysis"
author: "Michelle Mohr"
date: "2023-04-03"
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Introduction
Data from the North Pole Environmental Observatory Bottle Chemistry with measurements of various variables extracted from sea ice platforms. 

This data was downloaded on April 3, 2023 from [this link](https://arcticdata.io/catalog/view/doi:10.18739/A25T3FZ8X)

### Read data 

#### Set up

```{r library}
library(readr)
```


```{r read data, warning=FALSE, message=FALSE}
bg_chem <- read_csv("data/BGchem2008data.csv")
```

#### Data exploration 
```{r data expl, eval=FALSE, echo=FALSE}
## column names 
colnames(bg_chem)


## structure of my data
str(bg_chem)

##summary stats
summary(bg_chem)

## Open data in a tab
View(bg_chem)

## unique values in a column 
unique(bg_chem$Date)
```


## Analysis 

### Calculate summary statistics 
```{r mean calculation}
nitrate <- mean(bg_chem$NO3)
nitrite <- mean(bg_chem$NO2)
amm <- mean(bg_chem$NH4)
phos <- mean(bg_chem$P)

```

### Calculate mean Redfield ratio 

Redfield ratio 16:1
```{r ratio calc}

ratio <- (amm + nitrate + nitrite)/phos
```

### Plot Redfield ratio

```{r plot}
plot(bg_chem$P, bg_chem$NO3 + bg_chem$NO2 + bg_chem$NH4)
```


## Conclusion 

The Redfield ratio is `r round(ratio)`